Concurrent Programming in Harmony

### 2021-12-11 book/paper.tex
Latest

Harmony, too, has a Hello World program.
\autoref{fig:helloworld} shows the program and the corresponding output.
(The code for examples in this book can be found in the \texttt{code} folder under
the name listed in the caption of the example.)
The output is a \emph{Deterministic State Machine} (DFA).  The green circle
represents the initial state and the double circle represents the final
state.  There is one \emph{transition}, labeled with the string
``hello world''.  The DFA describes (or \emph{recognizes}) all possible
outputs that the program can generate.  In this case, there is only one.

\index{choose operator}%
But programs can usually have more than one execution and produce multiple
different outputs as a result.  This is usually as a result of different
inputs, but Harmony programs do not have inputs. Instead,
\autoref{fig:choice} demonstrates
\emph{nondetermistic choice} in Harmony programs.  In this case, the
program chooses to either print ``hello'' or ``world''.  The corresponding
DFA captures both possibilities.
You can think of the \texttt{choose} operator as enumerating all possible
inputs to the program.

\autoref{fig:hello4} shows a program that has an infinite number of outputs
by using a loop.  Harmony usually requires that any program must be able
to terminate, so the loop is conditioned on a nondeterministic choice between
\texttt{False} and \texttt{True}.  The possible outputs consists of zero or
more copies of the string ``hello world''.  Note that this single state DFA
(where the initial state and the final state happend to be the same)
captures an infinite number of possible outputs.

@@

``Running'' this Harmony program will try all possible executions, which
includes all possible values for $x$.  Try it out (here \texttt{\$}
represents a shell prompt):

@@

(For this to work, make sure \texttt{harmony} is in your command shell's search path.)
The $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, Harmony reports the size of the
``state graph'' (13 states in this case).  If not, Harmony also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

As with critical sections, when testing a concurrent data structure
we need a specification.
\autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behavior}.
%
\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to \texttt{NOPS}
queue operations.
It maintains two queues:
\begin{itemize}
\item[] \textit{implq}: the queue of the implementation;
\item[] \textit{specq}: the queue of the specification
\end{itemize}
For each operation, the code first decides whether to
perform a \texttt{get} or \texttt{put} operation.
In the case of a \texttt{put} operation, the code also
decides which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of \texttt{get}, the results of the operation
on both the implementation and specification are checked against one another.

@@

The next step is to test if the queue implementation
meets the \emph{concurrent} queue specification or not.
\autoref{fig:queues}(b) shows the concurrent version of
\autoref{fig:queues}(a)---it simply makes all operations
(except instantiation itself) atomic.

But we have not yet shown what the
concurrent queue specification is.

@@

\section*{Exercises}
\begin{problems}
\item Write a sequential specification for \autoref{fig:linkedlist}.
(Hint: it implements a set of integers.)
Write a Harmony program that checks if \autoref{fig:linkedlist} meets
the sequential specification.
\item Extend \autoref{fig:linkedlist} with linearization points and
check if the implementation is linearizable.
\item Check if the extended queue implementation of \autoref{ex:qremove}
is linearizable.
\end{problems}

@@

\begin{figure}
\begin{code}
\harmonysource{queuebug}
\end{code}
\caption{\harmonylink{code/queuebug.hny} A buggy queue implementation}
\label{fig:queuebug}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/queuebug1.pdf}
\end{center}
\caption{\harmonyref{queuebug.html}{Harmony output of running \autoref{fig:qtest3} against
\autoref{fig:queuebug}}}
\label{fig:queuebug1}
\end{figure}

@@

\autoref{fig:queuebug} contains an attempt at a queue implementation
where the queue is implemented by a linked list, with the first node
being a \texttt{dummy} node to prevent data races.  Each node in the
list contains a lock.  The \texttt{put}() method walks the list until
it gets to the last node, each time acquiring the lock to access the
node's fields.  When \texttt{put}() gets to the last node in the list,
it appends a new one.  The \texttt{get}() method locks the first (dummy)
node, removes the second from the list and frees it.  The method returns
the value from the removed node.

Let us run the code through some of the test programs in the last
chapter.  Harmony does not detect any issues with the sequential test
in \autoref{fig:qtestseq}.
(Run this using the \texttt{-m} flag like this:
\texttt{harmony -m queue=queuebug code/qtestseq.hny})
The concurrent \texttt{put} test of
\autoref{fig:qtest1} as well as the concurrent \texttt{get} test of
\autoref{fig:qtest2} also find no problems.  However, when we run
the new queue code through the test in \autoref{fig:qtest3}, Harmony
finds a problem.

\autoref{fig:queuebug1} shows the Harmony output of running
the test in \autoref{fig:qtest3} against the queue code
in \autoref{fig:queuebug}.
There is quite a bit of information in the Harmony output, and it's
important to learn to navigate through it.
Let's start with looking at the red text.  Harmony found a safety
violation (something bad happened during one of the possible executions),
and in particular \texttt{putter}(1) (thread T2) was trying to dereference
\texttt{alloc}.\textit{pool}[0].\texttt{lock} in turn 5.

@@

Next \texttt{putter}(1) (thread T2) takes a turn.  It gets to line 10 in
\autoref{fig:queuebug} where it's trying to acquire
$q$$\rightarrow$\texttt{lock}.  If we look at the bottom right,
we see that $q$ points to \textit{pool}[0], the node that has value 0
in it.  Again, so far so good.

@@

>>>>

Harmony, too, has a Hello World program.
\autoref{fig:helloworld} shows the program and the corresponding output.
From the command line, you can run it as follows:
\begin{code}
\begin{verbatim}
$ harmony -o hello1.png code/hello1.hny
\end{verbatim}
\end{code}
For this to work, make sure \texttt{harmony} is in your command shell's
search path.
The code for examples in this book can be found in the \texttt{code}
folder under the name listed in the caption of the example.
In this case, the file \texttt{code/hello1.hny} contains the
code in \autoref{fig:helloworld}.
The \texttt{-o hello1.png} arguments tell harmony to write the output
in the file \texttt{hello1.png}
The output is a \emph{Deterministic State Machine} (DFA).  The green circle
represents the initial state and the double circle represents the final
state.  There is one \emph{transition}, labeled with the string
``hello world''.  The DFA describes (or \emph{recognizes}) all possible
outputs that the program can generate.  In this case, there is only one.

\index{choose operator}%
But programs can usually have more than one execution and produce multiple
different outputs as a result.  This is usually as a result of different
inputs, but Harmony programs do not have inputs. Instead,
\autoref{fig:choice} demonstrates
\emph{nondetermistic choice} in Harmony programs.  In this case, the
program chooses to print either ``hello'' or ``world''.  The corresponding
DFA captures both possibilities.
You can think of the \texttt{choose} operator as enumerating all possible
inputs to the program.

\autoref{fig:hello4} shows a program that has an infinite number of outputs
by using a loop.  Harmony usually requires that any program must be able
to terminate, so the loop is conditioned on a nondeterministic choice between
\texttt{False} and \texttt{True}.  The possible outputs consist of zero or
more copies of the string ``hello world''.  Note that this single state DFA
(where the initial state and the final state happend to be the same)
captures an infinite number of possible outputs.

@@

Running this Harmony program will try all possible executions, which
includes all possible values for $x$.  Try it out (here \texttt{\$}
represents a shell prompt):

@@

The $\textbf{assert}$ statement checks that the output is
correct.  If the program is correct, Harmony reports the size of the
``state graph'' (13 states in this case).  If not, Harmony also
reports what went wrong, typically by displaying a summary of an execution in
which something went wrong.

@@

As with critical sections, when testing a concurrent data structure
we need a specification.
\autoref{fig:queues}(a) shows a sequential specification of a
queue in Harmony.
First we can check if the queue implementation
in \autoref{fig:queueconc} meets the sequential
queue specification in \autoref{fig:queues}(a).
To check if the queue implementation meets the specification,
we need to see if any sequence of queue operations in the
implementation matches a corresponding sequence in the
specification.
We say that the implementation and the specification
have the same \emph{behavior}.
%
\autoref{fig:qtestseq} presents a test program that does
exactly this, for sequences of up to \texttt{NOPS}
queue operations.
It maintains two queues:
\begin{itemize}
\item[] \textit{implq}: the queue of the implementation;
\item[] \textit{specq}: the queue of the specification
\end{itemize}
For each operation, the code first chooses whether to
perform a \texttt{get} or \texttt{put} operation.
In the case of a \texttt{put} operation, the code also
chooses which value to append to the queue.
All operations are performed on both the queue implementation
and the queue specification.
In the case of \texttt{get}, the results of the operation
on both the implementation and specification are checked against one another.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{../code/qtestpar.png}
\end{center}
\begin{code}
\harmonysource{qtestpar}
\end{code}
\caption{\harmonylink{code/qtestpar.hny} Concurrent queue test.  The
behavior DFA is for $\mathtt{NOPS} = 2$.}
\label{fig:qtestpar}
\end{figure}

\begin{figure}
\begin{code}
\harmonysource{queueseq}
\end{code}
\caption{\harmonylink{code/queueseq.hny} Sequential but not a concurrent
queue implementation}
\label{fig:queuebug}
\end{figure}

The next step is to test if the queue implementation
meets the \emph{concurrent} queue specification or not.
\autoref{fig:queues}(b) shows the concurrent queue specification.
It is similar to the sequential specification in \autoref{fig:queues}(a)
but makes all operations (except instantiation itself) atomic.
Testing the implementation of a concurrent queue specification is
trickier than testing the implementation of a sequential one because
there are many more scenarios to check.

We would like a way that systematically compares behaviors of the
concurrent queue implementation with behaviors of the concurrent
queue specification.  But we cannot do this by composing the specification
and the implementation and simply run the same test operations on both as we
did before---concurrency make the operations non-determistic and thus
the specification and implementation of a single execution might produce
different results, even if both are correct.
Instead, we will create a test program that tries various concurrent
combinations of queue operations, and run it twice: once against the
specification of the concurrent queue and once against the
implementation.  We will then check if the behaviors obtained from
running the implementation is a subset of the behaviors obtained
from the specification.

We will start with a test program that tries concurrent combinations
of various queue operations.
\autoref{fig:qtestpar} shows the test program.  It starts
\texttt{NOPS} threads and each choose to do either a \texttt{put}
or a \texttt{get} operation.  In case of a \texttt{put} operation,
the thread enqueues its own name.  In order to output the behaviors,
each thread prints what operation it is about to perform,
and afterwards it prints that the operation has
completed (including the return value if any).
The figure shows the deterministic finite automaton when the test
program is run against the specification in the case
$\texttt{NOPS} = 2$---for $\texttt{NOPS} = 4$ it would be much
too large to print here.  Since there are no cycles in the DFA, you
can follow some paths through this DFA and see that they are valid
interleavings of the threads.

If you run the same test program against the implementation
of \autoref{fig:queueconc}, you will get the same output.  You can
try this for various $\texttt{NOPS}$, although it gets increasingly
harder to check by hand that the generated DFAs are the same as
\texttt{NOPS} increases.
Now run the test program against \autoref{fig:queueseq}, which is
clearly an incorrect implementation of the concurrent queue specification
because it contains no synchronization code.
However, Harmony does not immediately detect
any problems.  In fact, for $\mathtt{NOPS} = 2$ it even generates
the same set of behaviors.
This is because the test program only outputs the behaviors---it does
not check if they are correct.

Harmony does have a way to check the behavior of one program against
the behavior of another.  In particular, we want to check if the
behaviors of the implementations we have match the behavior of the
specification.  The following shows, for example, how to check the
\texttt{queueconc.hny} implementation on the command line:

\begin{code}
\begin{verbatim}
$ harmony -o queue4.hfa code/qtestpar.hny
$ harmony -B queue4.hfa -m queue=queueconc code/qtestpar.hny
\end{verbatim}
\end{code}

The first command runs the \texttt{code/qtestpar.hny} program (with the
default 4 threads)
and writes a representation of the output DFA in the file \texttt{queue4.hfa}.
The second command runs the same test program, but using the queue
implementation in the file \texttt{code/queueconc.hny}.  Moreover, it reads
the DFA in \texttt{queue4.hfa} to check if every behavior of the second
run of the test program is also a behavior of the first run.
You can try the same using the \texttt{code/queueseq.hny} implementation
and find that this implementation has behaviors that are not allowed by
the specification.

@@

\section*{Exercises}
\begin{problems}
\item Write a sequential specification for \autoref{fig:linkedlist}.
(Hint: it implements a set of integers.)
Write a Harmony program that checks if \autoref{fig:linkedlist} meets
the sequential specification.
\item Also create a concurrent specification and a concurrent test program.
Check if the behaviors of \autoref{fig:linkedlist} match those of the
specification.
% \item Extend \autoref{fig:linkedlist} with linearization points and
% check if the implementation is linearizable.
% \item Check if the extended queue implementation of \autoref{ex:qremove}
% is linearizable.
\end{problems}

@@

\begin{figure}
\begin{code}
\harmonysource{queuebroken}
\end{code}
\caption{\harmonylink{code/queuebroken.hny} Another buggy queue implementation}
\label{fig:queuebroken}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/queuebug1.pdf}
\end{center}
\caption{\harmonyref{queuebug.html}{Harmony output of running \autoref{fig:qtest3} against
\autoref{fig:queuebroken}}}
\label{fig:queuebug1}
\end{figure}

@@

\autoref{fig:queuebroken} contains an attempt at a queue implementation
where the queue is implemented by a linked list, with the first node
being a \texttt{dummy} node to prevent data races.  Each node in the
list contains a lock.  The \texttt{put}() method walks the list until
it gets to the last node, each time acquiring the lock to access the
node's fields.  When \texttt{put}() gets to the last node in the list,
it appends a new one.  The \texttt{get}() method locks the first (dummy)
node, removes the second from the list and frees it.  The method returns
the value from the removed node.

Let us run the code through some of the test programs in the last
chapter.  Harmony does not detect any issues with the sequential test
in \autoref{fig:qtestseq}.
(Run this using the \texttt{-m} flag like this:
\texttt{harmony -m queue=queuebroken code/qtestseq.hny})
The concurrent \texttt{put} test of
\autoref{fig:qtest1} as well as the concurrent \texttt{get} test of
\autoref{fig:qtest2} also find no problems.  However, when we run
the new queue code through the test in \autoref{fig:qtest3}, Harmony
finds a problem.

\autoref{fig:queuebug1} shows the Harmony output of running
the test in \autoref{fig:qtest3} against the queue code
in \autoref{fig:queuebroken}.
There is quite a bit of information in the Harmony output, and it's
important to learn to navigate through it.
Let's start with looking at the red text.  Harmony found a safety
violation (something bad happened during one of the possible executions),
and in particular \texttt{putter}(1) (thread T2) was trying to dereference
\texttt{alloc}.\textit{pool}[0].\texttt{lock} in turn 5.

@@

Next \texttt{putter}(1) (thread T2) takes a turn.  It gets to line 10 in
\autoref{fig:queuebroken} where it's trying to acquire
$q$$\rightarrow$\texttt{lock}.  If we look at the bottom right,
we see that $q$ points to \textit{pool}[0], the node that has value 0
in it.  Again, so far so good.

@@

\chapter{Directly checking linearizability}
